RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-03 manuscripts/manuscript.Rmd
Rewrite of simmap section, part II.

Of course writing out such a definition manually becomes tedious quickly. Because
these are just R commands, we can easily define a function that can loop over an
assignment like this for each edge, extracting the appropriate order, length and
state from an existing R object such as that provided in the `phytools` package.  
Likewise, it is straightforward to define a function that reads this data using
the `RNeXML` utilities and converts it back to the `phytools` package. The full
implementation of this mapping can be seen in the `simmap_to_nexml()` and the
`nexml_to_simmap()` functions provided in the `RNeXML` package.

As the code indicates, the key step is simply to define the data in meta elements. In 
so doing, we have defined a custom namespace, `simmap`, to hold our variables.  This
allows us to provide a URL with more detailed descriptions of what each of these 
elements mean:

@@

At that URL we have posted a simple description of each term.

Using this convention we can generate NeXML files containing `simmap`
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a `nexml` object containing
a `simmap` reconstruction into R:

```{r}
data(simmap_ex)
```

The `get_trees()` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the `simmap` reconstruction data
and returns includes this in a `maps` element of the `ape::phylo` object,
for use with other `phytools` functions.

```{r}
phy <- nexml_to_simmap(simmap_ex)
```

We can then use various functions from `phytools` designed for `simmap`
objects [@Revell_2012], such as the plotting function:

```{r Figure1, fig.cap="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML."}
library("phytools")
plotSimmap(phy)
```

Likewise, we can convert the object back in the NeXML format and write
it out to file to be read by other users.

```{r}
nex <- simmap_to_nexml(phy) 
nexml_write(nex, "simmap.xml")
```

Though other NeXML parsers (for instance, for Perl or Python) have
not been written explicitly to express `simmap` data, those parsers will
nonetheless be able to successfully parse this file and expose the `simmap`
data to the user.

>>>>

The example uses a prefix, `simmap`, to group the newly introduced
metadata properties in a vocabulary, for which the `add_namespace()`
method can be used to give a URI as an identifier:

@@

As we illustrate here, even if the URI does not resolve to a fully
machine-readable definition of the terms and their semantics, it can
still be used to provide at least a human-readable informal definition
of the terms, and where to find such documentation _is_
machine-readable in the resulting NeXML document. Aside from the fact
that the above representation does not modify the NeXML format, and
should thus be readable -- even if not necessarily interpretable as a
stochastic character mapping -- by any compliant NeXML parser, the
direct and machine-readable link to metadata documentation contrasts
with metadata or format extensions that are undocumented, or which do
not directly link to documentation.

Although writing out metadata annotations as in the above proposal may
seem tedious, by virtue of `RNeXML`'s API they can be expressed as a
series of R commands, and it is therefore straightforward to define
functions that encapsulate the details behind a parameter
signature. To demonstrate this, we implemented two functions in
`RNeXML`, `simmap_to_nexml()` and `nexml_to_simmap()`, that map
bidirectionally between NeXML files with `simmap` metadata according
to the above proposal, and the `simmap` data representation in R
devised by @Revell_2012 in the R package `phytools`. The latter uses
an extension of the `ape::phylo` class. As shown in more detail in the
`RNeXML` package vignette, this allows generating NeXML files
containing `simmap` data, reading those files into R, converting them
into the in-memory representation understood by `phytools`, and using
the various functions from `phytools` designed for `simmap` objects
[@Revell_2012], such as `plotSimmap()` for plotting.
